---
title: "ST4060/ST6040 -- Section 3: Introduction to Machine Learning"
author: "Eric Wolsztynski, Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(ISLR)
knitr::opts_chunk$set(echo = FALSE)
```

<!-- --------------------------------------------------------------------- -->
## Classification: logistic regression

Recall from the notes: logistic regression is a particular case of GLM used
for classification. Here we use the `Default` dataset (library `ISLR`), where
the response `default` indicates whether a credit card customer defaulted.

**Exercise 1:** fit a logistic regression of `default` on `balance` using
`glm()` with `family="binomial"`, and interpret the sign and significance of
the coefficient on `balance`.

```{r logreg-1, exercise = TRUE, exercise.lines = 8}

```

```{r logreg-1-solution}
glmo = glm(default~balance, data=Default, family="binomial")
summary(glmo)
# positive, highly significant coefficient on balance:
# higher balance is associated with higher probability of default
```

**Exercise 2:** extend the model of Exercise 1 to also include `income` and
`student`. Has the effect of `balance` changed much? What do you make of the
coefficient on `student`?

```{r logreg-2, exercise = TRUE, exercise.lines = 8}

```

```{r logreg-2-solution}
glmo = glm(default~balance+income+student, data=Default, family="binomial")
summary(glmo)
```

**Exercise 3:** split `Default` into a training set (70% of observations) and
a test set (the remaining 30%). Fit the model of Exercise 2 on the training
set only, then use it to predict `default` on the test set (recall
`predict(..., type="response")` returns estimated probabilities, which you
need to threshold at 0.5 to obtain class predictions). Compute the confusion
matrix and the test error rate.

```{r logreg-3, exercise = TRUE, exercise.lines = 15}

```

```{r logreg-3-hint}
set.seed(1)
itrain = sample(1:nrow(Default), round(.7*nrow(Default)))
train = Default[itrain,]
test  = Default[-itrain,]

glmo = glm(default~balance+income+student, data=train, family="binomial")
probs = predict(glmo, newdata=test, type="response")
preds = ifelse(probs>0.5, "Yes", "No")

table(predicted=preds, actual=test$default)
mean(preds != test$default) # test error rate
```

**Exercise 4:** the *null* classifier that always predicts "No" (i.e. never
predicts a default) can already achieve a low error rate here, simply because
defaults are rare in this dataset. Compute this null error rate on the test
set, and compare it with the error rate obtained in Exercise 3. What does this
tell you about the limits of overall error rate as a performance measure for
imbalanced classification problems?

```{r logreg-4, exercise = TRUE, exercise.lines = 8}

```

<!-- --------------------------------------------------------------------- -->
## Principal Component Analysis

Recall from the notes: PCA re-expresses a set of (possibly correlated)
variables as a smaller set of uncorrelated components, ordered by the amount
of variance they explain. Here we first review the mechanics of PCA against
its underlying eigen-decomposition, then apply it to the `iris` dataset.

**Exercise 1:** generate a 2D correlated dataset with
```{r pca-setup, exercise = FALSE, echo = TRUE, eval = FALSE}
set.seed(1)
n <- 100
X <- matrix(rnorm(n*2), ncol=2)
X[,2] <- 0.8*X[,1] + 0.6*X[,2]
```
Compute the eigenvalues of the correlation matrix of `X` (using `eigen()`),
and compare them with the squared standard deviations returned by a scaled
`prcomp()` on `X`. Are they the same, up to numerical precision?

```{r pca-1, exercise = TRUE, exercise.lines = 10}

```

```{r pca-1-solution}
set.seed(1)
n <- 100
X <- matrix(rnorm(n*2), ncol=2)
X[,2] <- 0.8*X[,1] + 0.6*X[,2]

Sig <- cor(X)
ev  <- eigen(Sig, symmetric=TRUE)
pca <- prcomp(X, scale=TRUE, center=TRUE)

ev$values         # eigenvalues of the correlation matrix
pca$sdev^2        # squared standard deviations from prcomp
# these match up to numerical (floating-point) precision
```

**Exercise 2:** perform PCA of the `iris` dataset (its first 4, numeric,
columns only), setting both `center` and `scale` to `TRUE`. Using
`summary()`, determine how many components are needed to capture at least
90% of the total variance.

```{r pca-2, exercise = TRUE, exercise.lines = 8}

```

```{r pca-2-solution}
pca.x = prcomp(iris[,1:4], center=TRUE, scale=TRUE)
summary(pca.x)
# PC1 alone explains ~73% of variance; PC1+PC2 together explain ~96%,
# so 2 components are already enough to pass the 90% threshold
```

**Exercise 3:** create a scatterplot of the first two principal components
from Exercise 2, colouring points by `iris$Species`. Compare this with a
scatterplot of the two original variables `Sepal.Length` and `Sepal.Width`,
also coloured by species. Which of the two plots separates the three species
more clearly?

```{r pca-3, exercise = TRUE, exercise.lines = 10}

```

```{r pca-3-hint}
pca.x = prcomp(iris[,1:4], center=TRUE, scale=TRUE)
par(mfrow=c(1,2), pch=20)
plot(iris$Sepal.Length, iris$Sepal.Width, col=iris$Species,
     xlab="Sepal.Length", ylab="Sepal.Width", main="Original variables")
plot(pca.x$x[,1], pca.x$x[,2], col=iris$Species,
     xlab="PC1", ylab="PC2", main="First two PCs")
# The PC1-PC2 plot separates the three species much more clearly than the
# original Sepal.Length/Sepal.Width plot, particularly versicolor vs
# virginica -- consistent with PC1 pooling information from all 4 original
# (correlated) variables rather than just 2 of them.
```
