---
title: "ST4060/ST6040 -- Section 3: Introduction to Machine Learning"
author: "Eric Wolsztynski, Statistics, School of Mathematical Sciences, University College Cork"
date: "2026-27"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(shiny)
library(ISLR)
knitr::opts_chunk$set(echo = FALSE)
```

<!-- --------------------------------------------------------------------- -->
## Classification: logistic regression

Recall from the notes: logistic regression is a particular case of GLM used
for classification. Here we use the `Default` dataset (library `ISLR`), where
the response `default` indicates whether a credit card customer defaulted.

**Exercise 1:** fit a logistic regression of `default` on `balance` using
`glm()` with `family="binomial"`, and interpret the sign and significance of
the coefficient on `balance`.

```{r logreg-1, exercise = TRUE, exercise.lines = 8}

```

```{r logreg-1-solution}
glmo = glm(default~balance, data=Default, family="binomial")
summary(glmo)
# positive, highly significant coefficient on balance:
# higher balance is associated with higher probability of default
```

**Exercise 2:** extend the model of Exercise 1 to also include `income` and
`student`. Has the effect of `balance` changed much? What do you make of the
coefficient on `student`?

```{r logreg-2, exercise = TRUE, exercise.lines = 8}

```

```{r logreg-2-solution}
glmo = glm(default~balance+income+student, data=Default, family="binomial")
summary(glmo)
```

**Exercise 3:** split `Default` into a training set (70% of observations) and
a test set (the remaining 30%). Fit the model of Exercise 2 on the training
set only, then use it to predict `default` on the test set (recall
`predict(..., type="response")` returns estimated probabilities, which you
need to threshold at 0.5 to obtain class predictions). Compute the confusion
matrix and the test error rate.

```{r logreg-3, exercise = TRUE, exercise.lines = 15}

```

```{r logreg-3-hint}
set.seed(1)
itrain = sample(1:nrow(Default), round(.7*nrow(Default)))
train = Default[itrain,]
test  = Default[-itrain,]

glmo = glm(default~balance+income+student, data=train, family="binomial")
probs = predict(glmo, newdata=test, type="response")
preds = ifelse(probs>0.5, "Yes", "No")

table(predicted=preds, actual=test$default)
mean(preds != test$default) # test error rate
```

**Exercise 4:** the *null* classifier that always predicts "No" (i.e. never
predicts a default) can already achieve a low error rate here, simply because
defaults are rare in this dataset. Compute this null error rate on the test
set, and compare it with the error rate obtained in Exercise 3. What does this
tell you about the limits of overall error rate as a performance measure for
imbalanced classification problems?

```{r logreg-4, exercise = TRUE, exercise.lines = 8}

```
